---
title: 'Exercise 1: Variance Estimation'
author: "Joan Salvà"
date: "October 2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: Advanced Methods for Regression and Classification
---


We load the `College` dataset as for the last exercise, log-transform the target variable, split into train/test and define the RMSE evaluation.

```{r, results='hold'}
data(College, package='ISLR')

library(data.table)
df <- setDT(College)

# Make this example reproducible
set.seed(1)

# Do not use Accept and Enroll
out_of_model <- c('Accept', 'Enroll', 'Apps')

# randomly split into training and test data:
n <- nrow(College)
index <- sample(1:n, round(n *2/3))

df$y <- log(df$Apps)

train  <- df[index, !..out_of_model]
test   <- df[!index, !..out_of_model]

evaluate_on_set <- function(mod, set) {
  
  s <- sqrt(
    mean((predict(mod, set) - set$y)^2)
  )
  return(s)
}
```

### Exercise 1

```{r}
full_model <- lm(y ~ ., data=train)
summary(full_model)
plot(full_model)
```

```{r}
library(cvTools)

cv_fit <- cvFit(full_model, data = train, y = train$y, K = 5, cost = rmspe, R = 100, seed=1)
cv_fit
plot(cv_fit)
```

What do we conclude? 

### Exercise 2: Best subset regression

```{r, results = 'hide'}
library(leaps)

reg <- regsubsets(y ~ ., data = train, nbest = 3, nvmax = 10)
summary(reg)
plot(reg)
```


We can see that no model has a BIC below -600. Therefore, the best model will be the simplest one in that BIC range. That is the model with subset size = 8 with features *PrivateYes*, *F.Undergrad*, *Room.Board*, *Books*, *PhD*, *S.F.Ratio*, *Expend*, *Grad.Rate*. However, the table above is unclear and we would have to find a better way to plot the results.

```{r}
x <- summary(reg)

str(x)
index <- numeric(0)
for (i in 1:10) {
  index <- c(index, rep(i, 3))
}
index

plot(index, x$bic, xlab='Size of subset', ylab = 'BIC', main = 'BIC by number of regressors')

```
The best model is found with size = 9. It has the subset of variables: *PrivateYes*, *F.Undergrad*, *Outstate*, *Room.Board*, *Books*, *PhD*, *S.F.Ratio*, *Expend*, *Grad.Rate*, in addition to the intercept.

Let's test this reduced linear model against the full model in (1) using the `cvFit` function.

```{r}
reduced_mod <- lm(y ~ Private + F.Undergrad + Outstate + Room.Board + Books + PhD + S.F.Ratio + Expend + Grad.Rate, data = train)

summary(reduced_mod)
```
In the reduced model, all variables show strong significance. That is not the case of the full model, as we had seen before.

```{r, results='hold'}

cv_fit_reduced <- cvFit(reduced_mod, data = train, y = train$y, K = 5, cost = rmspe, R = 100, seed=1)
print('Reduced model CV results: ')
cv_fit_reduced
print('Full model CV results: ')
cv_fit


plot(cv_fit_reduced)
plot(cv_fit)

evaluations <- c(
  round(evaluate_on_set(full_model, train), 3),
  round(evaluate_on_set(full_model, test), 3),
  round(evaluate_on_set(reduced_mod, train), 3),
  round(evaluate_on_set(reduced_mod, test), 3)
)

cv_list <- c(
  round(cv_fit$cv, 3), ' ' , round(cv_fit_reduced$cv, 3), ' ' 
)

rsqrt <- c(round(summary(full_model)$adj.r.squared, 3), '', round(summary(reduced_mod)$adj.r.squared, 3), '')


table <- data.frame(
  Dataset=c('train', 'test', 'train', 'test'),
  Model=rep(c('full', 'reduced'), each=2),
  RMSE=evaluations, 
  CVFit=cv_list, 
  AdjRSqrt=rsqrt
)

knitr::kable(table, caption = "Error table")
```
Prediction error seems to be a bit smaller in the reduced model. This one presents errors with lower variance over the 100 iterations of the cross-validation.

However, the RMSE and $R^2$ values seem to be slightly worse in the reduced model.

### Exercise 3: Principal component regression (PCR)

```{r}

model_pcr <- plsr(y~ ., data = df[, !..out_of_model], susbet = train, scale = TRUE, validation="CV", segments=10, segment.type="random")

summary(model_pcr)

plot(model_pcr, plottype = "validation", val.type = "RMSEP", legend = "topright")

```
The model with 12 components seems optimal because the RMSE does not get better when more components are added.

```{r}
number_of_components = 13
plot(model_pcr, ncomp=number_of_components, asp=1, line=TRUE)

pred_train <- predict(model_pcr, newdata=train, ncomp=number_of_components)
pred_test <- predict(model_pcr, newdata=test, ncomp=number_of_components)
rmse_train <- sqrt(mean((train$y - pred_train)^2))
rmse_test <- sqrt(mean((test$y - pred_test)^2))

print(paste('RMSE (train):', rmse_train))
```


```{r}

predplot(model_pcr, ncomp = number_of_components, which = c('test'), newdata = test)
abline(a = 0, b = 1)
print(paste('RMSE (test):', rmse_test))

```

### Exercise 4: Partial least squares regression (PLS)


```{r}
library(pls)

model_plsr <- plsr(y~ ., data = df[, !..out_of_model], susbet = train, scale = TRUE, validation="CV", segments=10, segment.type="random")

summary(model_plsr)

plot(model_plsr, plottype = "validation", val.type = "RMSEP", legend = "topright")

```
The model with 5 components seems optimal because the RMSE does not get better when more components are added.

```{r}
number_of_components = 5
plot(model_plsr, ncomp=number_of_components, asp=1, line=TRUE)

pred_train <- predict(model_plsr, newdata=train, ncomp=number_of_components)
pred_test <- predict(model_plsr, newdata=test, ncomp=number_of_components)
rmse_train_plsr <- sqrt(mean((train$y - pred_train)^2))
rmse_test_plsr <- sqrt(mean((test$y - pred_test)^2))

print(paste('RMSE (train):', rmse_train_plsr))
```


```{r}

predplot(model_plsr, ncomp = number_of_components, which = c('test'), newdata = test)
abline(a = 0, b = 1)
print(paste('RMSE (test):', rmse_test_plsr))

```
*PCR vs PLSR comparison*
```{r}
evaluation <- c(
  rmse_train, rmse_test, rmse_train_plsr, rmse_test_plsr
)

table <- data.frame(
  Dataset=c('train', 'test', 'train', 'test'),
  Model=rep(c('PCR', 'PLS'), each=2),
  optimal_number_of_components = c(13,13,5,5),
  RMSE=evaluations
)
colnames(table) <- c('Dataset', 'Model', 'Optimal N. Components', 'RMSE')

knitr::kable(table, caption = "Error table")
```

PCR provides better RMSE values both in the test and train datasets. However, it uses more components. 

### Exercise 5: PCR by hand

```{r}
X <- scale(model.matrix(train$y ~ ., train), center = TRUE, scale = TRUE)
X[, 1] <- numeric(length = nrow(X))

principal_components <- princomp(X)

q <- 13

Z_train <- principal_components$scores[, 1:q]

theta <- solve(t(Z_train) %*% Z_train) %*% t(Z_train) %*% scale(train$y, center = TRUE, scale = TRUE)

X_test <- scale(model.matrix(test$y ~ ., test), scale = TRUE, center = TRUE)
X_test[, 1] <- numeric(length = nrow(X_test))
Z_test <- X_test %*% principal_components$loadings


prediction <- Z_test[, 1:q] %*% theta
print(prediction)


plot(prediction, scale(test$y, center = TRUE, scale = TRUE))

rmse_manual_pcr <- sqrt(mean((prediction - scale(test$y, center = TRUE, scale = TRUE))^2))
print(rmse_manual_pcr)
abline(a = 0, b = 1)

```
```{r}
model_pcr$coefficients[, , q]

beta_manual <- principal_components$loadings[, 1:q] %*% theta
print(beta_manual)
```

</div>
</pre>
