---
title: 'Assignment 4: Sample distribution and Central Limit Theorem'
author: "Joan Salvà"
date: "November 2022"
output:
  pdf_document: default
  html_document:
  df_print: paged
subtitle: Statistical Simulation and Computerintensive Methods
---


# Task 1
Consider the 12 sample data points: 4.94 5.06 4.53 5.07 4.99 5.16 4.38 4.43 4.93 4.72 4.92 4.96
```{r}
data <- c(4.94, 5.06, 4.53, 5.07, 4.99, 5.16, 4.38, 4.43, 4.93, 4.72, 4.92, 4.96)
n <- length(data)
```

1. How many possible bootstrap samples are there, if each bootstrap sample has the same size as the original?
*Answer*: Since bootstrapping does not pay attention to the order and allows repetition:
$$\text{Number of boostrap samples} = {2n + 1\choose n} = {25 \choose 12} = 5200300$$

2. Compute the mean and the median of the original sample.
```{r}
original_mean <- mean(data)
original_median <- median(data)
print(paste('Mean:', original_mean))
print(paste('Median:', original_median))
```

3. Create 2000 bootstrap samples and compute their means.
```{r}
matrikelnummer <- 12223411
set.seed(matrikelnummer)
n_samples <- 2000
samples <- replicate(2000, sample(data, size=n, replace=TRUE))
means <- numeric(2000)
for(i in 1:2000){
  means[i] <- mean(samples[,i]) 
}
```

3.1, 3.2, 3.3 Compute the mean on the first 20, 200, and 2000 bootstrap means.
```{r}
print(paste('First 20: ', mean(means[1:20])))
print(paste('First 200: ', mean(means[1:200])))
print(paste('First 2000: ', mean(means[1:200])))
```

3.4 Visualise the distribution all the different bootstrap means to the sample mean. Does the Central Limit Theorem kick in?
```{r}
hist(means, breaks = 50)
abline(v=original_mean, col='orange')
```
*Answer*: Yes, the Central Limit Theorem kicks in because the means seem to follow a Normal Distribution around the original mean.

5. Based on the three different bootstrap sample lengths in 3. compute the corresponding 0.025 and 0.975 quantiles. Compare the three resulting intervals against each other and the "true" confidence interval of the mean under the assumption of normality. (Use for example the function t.test to obtain the 95% percent CI based on asympotic considerations for the mean.)
```{r}
a <- .025
b <- 1-a
print(paste('Conf.  Int. for means (20):', quantile(means[1:20], a), quantile(means[1:20], b)))

print(paste('Conf.  Int. for means (200):', quantile(means[1:200], a), quantile(means[1:200], b)))

print(paste('Conf.  Int. for means (2000):', quantile(means[1:2000], a), quantile(means[1:2000], b)))



true_confidence_interval <- t.test(data, conf.level = 0.95)$conf.int
true_confidence_interval

```
*Answer*: When computing the confidence intervals from the increasing sized mean vectores, the interval seems to converge to $[4.7, 4.975]$. However, the confidence interval coming from the T test suggests a slighly bigger interval: $[4.67, 5]$.


4. Create 2000 bootstrap samples and compute their medians.
```{r}
set.seed(matrikelnummer)
n_samples <- 2000
samples <- replicate(2000, sample(data, size=n, replace=TRUE))
medians <- numeric(2000)
for(i in 1:2000){
  medians[i] <- median(samples[,i]) 
}
```

4.1, 4.2, 4.3 Compute the mean on the first 20, 200, and 2000 bootstrap medians.
```{r}
print(paste('First 20: ', mean(medians[1:20])))
print(paste('First 200: ', mean(medians[1:200])))
print(paste('First 2000: ', mean(medians[1:200])))
```
4.4 Visualise the distribution all the different bootstrap medians to the sample median.
```{r}
hist(medians)
abline(v=original_median, col='orange')
```

4.5 Based on the three different bootstrap sample lengths in 3. compute the corresponding 0.025 and 0.975 quantiles. Compare the three resulting intervals against each other.
```{r}
a <- .025
b <- 1 - a
print(paste('Conf.  Int. for medians (20):', quantile(medians[1:20], a), quantile(medians[1:20], b)))

print(paste('Conf.  Int. for medians (200):', quantile(medians[1:200], a), quantile(medians[1:200], b)))

print(paste('Conf.  Int. for medians (2000):', quantile(medians[1:2000], a), quantile(medians[1:2000], b)))

```

# Task 2
We wish to explore the effect of outliers on the outcomes of Bootstrap Sampling.

1. Set your seed to 1234. And then sample 1960 points from a standard normal distribution to create the vector x.clean then sample 40 observations from uniform(4,5) and denote them as x.cont. The total data is x <- c(x.clean,x.cont). After creating the sample set your seed to your immatriculation number.

```{r}
set.seed(1234)
x.clean <- rnorm(n = 1960)
x.cont <- runif(40, min = 4, max = 5)

x <- c(x.clean, x.cont)
set.seed(matrikelnummer)

```

2. Estimate the median, the mean and the trimmed mean with alpha = 0.05 for x and x.clean.

```{r}
table <- data.frame(
  sample=c('X', 'X.clean'),
  Mean=c(mean(x), mean(x.clean)),
  Median=c(median(x), median(x.clean)),
  TrimmedMean=c(mean(x, trim=0.05), mean(x.clean, trim=0.05))
)

knitr::kable(table, caption = "X vs X.clean")
```


3. Use nonparametric bootstrap (for x and x.clean) to calculate:
- The standard error
- The 95 percentile CI of all 3 estimators.
```{r}

m <- 2000
samples_x <- replicate(m, sample(x, size=length(x), replace=TRUE))
means_x <- numeric(m)
for (i in 1:m) {means_x[i] <- mean(samples_x[,i])}

samples_x_clean <- replicate(m, sample(x, size=length(x.clean), replace=TRUE))
means_x_clean <- numeric(length(samples_x_clean))
for (i in 1:m) {means_x_clean[i] <- mean(samples_x_clean[,i])}

standard_error_x <- sqrt(1 / (m - 1) * sum((means_x - mean(means_x))^2))

standard_error_x_clean <- sqrt(1 / (m - 1) * sum((means_x_clean - mean(means_x_clean))^2))

a <- 0.05
b <- 1 - a
ci_mean = c(
  quantile(means_x, a), quantile(means_x, b),
  quantile(means_x_clean, a), quantile(means_x_clean, b)
)

medians_x <- numeric(m)
for (i in 1:m) {means_x[i] <- median(samples_x[,i])}
medians_x_clean <- numeric(m)
for (i in 1:m) {means_x_clean[i] <- median(samples_x_clean[,i])}
trimmean_x <- numeric(m)
for (i in 1:m) {trimmean_x[i] <- mean(samples_x[,i], trim=0.05)}
trimmean_x_clean <- numeric(m)
for (i in 1:m) {trimmean_x_clean[i] <- mean(samples_x_clean[,i], trim=0.05)}

ci_median = c(
    quantile(medians_x, a), quantile(medians_x, b),
    quantile(medians_x_clean, a), quantile(medians_x_clean, b))

ci_trimmean = c(
  quantile(trimmean_x, a), quantile(trimmean_x, b),
  quantile(trimmean_x_clean, a), quantile(trimmean_x_clean, b))

table <- data.frame(
  sample=c('X', 'X.clean'),
  StandardError=c(standard_error_x, standard_error_x_clean),
  CIMeanLeft=c(ci_mean[1], ci_mean[3]),
  CIMeanRight=c(ci_mean[2], ci_mean[4]),
  CIMedianLeft=c(ci_median[1], ci_median[3]),
  CIMedianRight=c(ci_median[2], ci_median[4]),
  CITrimMeanLeft=c(ci_trimmean[1], ci_trimmean[3]),
  CITrimMeanRight=c(ci_trimmean[2], ci_trimmean[4])
)

knitr::kable(table, caption = "SE and Confidence intervals for different estimators")

```

4, 5. Use parametric bootstrap (based on x and x.clean) to calculate, for the mean and the trimmed mean:
- bias
- standard error
- 95 percentile CI
- bias corrected estimate

Compare and summarize your findings with tables and graphically.

```{r}
mean_estimate_x <- mean(x)
mean_estimate_x_clean <- mean(x.clean)
sd_estimate_x <- sd(x)
sd_estimate_x_clean <- sd(x.clean)
trimmean_estimate_x <- mean(x, trim=0.05)
trimmean_estimate_x_clean <- mean(x.clean, trim=0.05)

samples_x <- replicate(m, rnorm(length(x), mean=mean_estimate_x, sd=sd_estimate_x))
means_x <- numeric(m)
for (i in 1:m) {means_x[i] <- mean(samples_x[,i])}


samples_x_clean <- replicate(m, rnorm(length(x.clean), mean=mean_estimate_x_clean, sd=sd_estimate_x_clean))
means_x_clean <- numeric(length(samples_x_clean))
for (i in 1:m) {means_x_clean[i] <- mean(samples_x_clean[,i])}

standard_error_x <- sqrt(1 / (m - 1) * sum((means_x - mean(means_x))^2))

standard_error_x_clean <- sqrt(1 / (m - 1) * sum((means_x_clean - mean(means_x_clean))^2))

a <- 0.05
b <- 1 - a
ci_mean = c(
  quantile(means_x, a), quantile(means_x, b),
  quantile(means_x_clean, a), quantile(means_x_clean, b)
)

medians_x <- numeric(m)
for (i in 1:m) {means_x[i] <- median(samples_x[,i])}
medians_x_clean <- numeric(m)
for (i in 1:m) {means_x_clean[i] <- median(samples_x_clean[,i])}
trimmean_x <- numeric(m)
for (i in 1:m) {trimmean_x[i] <- mean(samples_x[,i], trim=0.05)}
trimmean_x_clean <- numeric(m)
for (i in 1:m) {trimmean_x_clean[i] <- mean(samples_x_clean[,i], trim=0.05)}

ci_median = c(
    quantile(medians_x, a), quantile(medians_x, b),
    quantile(medians_x_clean, a), quantile(medians_x_clean, b))

ci_trimmean = c(
  quantile(trimmean_x, a), quantile(trimmean_x, b),
  quantile(trimmean_x_clean, a), quantile(trimmean_x_clean, b))

bias_mean_x <- mean(means_x) - mean_estimate_x
bias_trim_mean_x <- mean(trimmean_x) - trimmean_estimate_x

bias_mean_x_clean <- mean(means_x_clean) - mean_estimate_x_clean
bias_trim_mean_x_clean <- mean(trimmean_x_clean) - trimmean_estimate_x_clean

table <- data.frame(
  sample=c('X', 'X.clean'),
  BiasMean=c(bias_mean_x, bias_mean_x_clean),
  BiasTrimMean=c(bias_trim_mean_x, bias_trim_mean_x_clean),
  StandardError=c(standard_error_x, standard_error_x_clean)
)

knitr::kable(table, caption = "SE and Bias for the estimates")

table_2 <- data.frame(
  sample=c('X', 'X.clean'),
  CIMeanLeft=c(ci_mean[1], ci_mean[3]),
  CIMeanRight=c(ci_mean[2], ci_mean[4]),
  CIMedianLeft=c(ci_median[1], ci_median[3]),
  CIMedianRight=c(ci_median[2], ci_median[4]),
  CITrimMeanLeft=c(ci_trimmean[1], ci_trimmean[3]),
  CITrimMeanRight=c(ci_trimmean[2], ci_trimmean[4])
)
knitr::kable(table_2, caption = "Confidence intervals for the estimates")

table <- data.frame(
  sample=c('X', 'X.clean'),
  MeanEstimator=c(mean_estimate_x, mean_estimate_x_clean),
  TrimMeanEstimator=c(trimmean_estimate_x, trimmean_estimate_x_clean),
  bias_corrected_mean=c(mean_estimate_x - bias_mean_x, mean_estimate_x_clean - bias_mean_x_clean),
  bias_corrected_trimmean=c(trimmean_estimate_x - bias_trim_mean_x, trimmean_estimate_x_clean - bias_trim_mean_x_clean)
)
names(table) <- c(
  'Sample', 'Mean Est.', 'Trimmed Mean Est.', 'Unbiased Mean Est.', 'Unbiased Trimmed Mean Est.'
)
knitr::kable(table, caption = "Estimators and Bias Corrected Estimators")
```

# Task 3
Based on the above tasks and your lecture materials, explain the methodology of bootstrapping for the construction of confidence intervals and parametric or non-parametric tests.

*Answer*: The main idea of bootstrapping is resampling a single dataset to create many simulated samples. This process allows you to calculate standard errors, construct confidence intervals, and perform hypothesis testing.

Let's go over the process of defining a confidence interval for some parameter $theta$ that describes a distribution and that we want to estimate. In Task 1 and 2, we used the Percentile Bootstrap CI method, that works as follows: resample $m$ times from the original sample and compute the estimator for each sample, giving $\hat{\theta}^i, i = 1, \dots, m$ estimates. Then use the percentils of these sample of size $m$ to define the confidence interval. Typically choose $p_1  = 0.05$ and $p_2 = 0.05$. 

Bootstrap tests are useful when the alternative hypothesis is not well specified. In cases where there is parametric alternative hypothesis, likelihood or Bayesian methods might be preferable. There are multiple tests that use bootstrapping resampling. A basic exaple is the following:

Let ${ x_{1},\ldots ,x_{n}}x_{1},\ldots ,x_{n}$ be a random sample from distribution F with sample mean ${ {\bar {x}}}{\bar {x}}$ and sample variance ${ \sigma _{x}^{2}}\sigma _{x}^{2}$. Let ${ y_{1},\ldots ,y_{m}}{ y_{1},\ldots ,y_{m}}$ be another, independent random sample from distribution G with mean ${ {\bar {y}}}{\bar {y}}$ and variance ${ \sigma _{y}^{2}}\sigma _{y}^{2}$

1. Calculate the test statistic ${ t={\frac {{\bar {x}}-{\bar {y}}}{\sqrt {\sigma _{x}^{2}/n+\sigma _{y}^{2}/m}}}}$
2. Create two new data sets whose values are ${ x_{i}'=x_{i}-{\bar {x}}+{\bar {z}}}$ and ${ y_{i}'=y_{i}-{\bar {y}}+{\bar {z}},}$ where ${ {\bar {z}}}$ is the mean of the combined sample.
3. Draw a random sample (${ x_{i}^{*}}$) of size $n$ with replacement from ${ x_{i}'}$ and another random sample (${ y_{i}^{*}}$) of size $m$ with replacement from ${ y_{i}'}$.
4. Calculate the test statistic ${ t^{*}={\frac {{\bar {x^{*}}}-{\bar {y^{*}}}}{\sqrt {\sigma _{x}^{*2}+v \sigma _{y}^{*2}/m}}}}$
5. Repeat 3 and 4 $B$ times (for example $B$ = 1000 times) to collect $B$ values of the test statistic.
6. Estimate the p-value as ${ p={\frac {\sum _{i=1}^{B}I\{t_{i}^{*}\geq t\}}{B}}}$ where ${ I({\text{condition}})=1}$ when condition is true and 0 otherwise.

</div></pre>
