---
title: "Exercise 3 - Monte Carlo Simulation of areas"
author: "Yannik Gaebel"
date: '2022-10-30'
output: pdf_document
---

```{r, echo = TRUE}
set.seed(12208157)
```

# Task 1

Plot the function

```{r}
x <- seq(-1,3,0.1)
g <- function(x) {exp(-x^3)}
exp <- dexp(x,rate=1)
plot(x,g(x),type='l')
lines(x,exp(x),type='l')
```

# Task 1.1

```{r}
# Calculate Monte Carlo Integral
b <- 6
X <- runif(100000, min=1, max=b)
cat(paste("Monte Carlo Integral:",mean(exp(-X^3)*(b-1))),"\n")

# Calculate Numerical Integral
cat(paste("Numerical Integral: "))
integrate(f = function(x){exp(-x^3)},lower = 1,upper = b)
```

Off by 0.002

# Task 1.2

What would be a good density for the simulation in that case?

Answer:

exponential distribution 

We change the formula in the following way so that the exponential distribution can be used

```{r}
X <- rexp(100000)
cat(paste("Monte Carlo Integral:",mean(exp(-(X+1)^3)/dexp(X))),"\n")

# Calculate Numerical Integral
cat(paste("Numerical Integral: "))
integrate(f = function(x){exp(-x^3)},1,Inf)
```

Off by 0.0003

# Task 1.3

Do you have an explanation why Monte Carlo integration agrees in 2. with integrate but not so much in 1.?

Answer:

Fläche welche öfter berechnet ist besser agbedeckt. Fläche die für Integral relevant ist wird am meisten "getestet"

# Task 2.1

```{r}
t <- seq(-pi,pi,0.01)
r <- function(t) {exp(cos(t))-2*cos(4*t)-sin(t/12)^5}
plot(t,r(t),type='l')
```

# Task 2.2

```{r}
r_x <- runif(1)
r_y <- runif(1)
lies
```

# Task 2.3

```{r}

```

# Task 2.4

```{r}

```

